---
title: "Pre-course preparation"
author: "Denis Rustand (denis.rustand@u-bordeaux.fr)"
date: "September 27, 2026"
output: pdf_document
subtitle: "Efficient Bayesian Analysis of Longitudinal and Survival Data Using R-INLA
  \\\nISCB 2026, Freiburg, Germany\n"
---

# Before the course

Running the code during the course is **optional**. If you want to run the code, please complete the following steps **before the course**. Installing packages can take time, so do not leave this for the last minute.

GITHUB : <https://github.com/DenisRustand/ISCB2026>

## R version

You need **R \>= 4.6**. Check with:

```{r check-r, eval=FALSE}
R.version.string
```

If your version is older, update from <https://cloud.r-project.org>.

## Package installation

```{r install, eval=FALSE}
install.packages("INLA",
  repos = c(getOption("repos"),
            INLA = "https://inla.r-inla-download.org/R/testing"))

install.packages(c("INLAjoint", "ggplot2", "ggpubr", "survival",
                   "mvtnorm", "PermAlgo", "JM",
                   "rmarkdown", "knitr", "sn"))
```

## Verify installation

Run the following to check that the core packages load and compute correctly:

```{r verify, eval=FALSE}
library(INLA)
library(INLAjoint)
cat("INLA version:", inla.version()$version, "\n")
cat("INLAjoint version:", as.character(packageVersion("INLAjoint")), "\n")

# Toy example: intercept of y ~ 1 with two observations (2 and 8) should be 5
m <- inla(y ~ 1, data = list(y = c(2, 8)),
          control.compute = list(config = TRUE))
cat("Intercept (expected 5):", round(m$summary.fixed["(Intercept)", "mean"], 1), "\n")
invisible(inla.posterior.sample(1, m))
cat("Posterior sampling: OK\n")
```

On Linux, if INLA throws a binary error, try switching binaries:

```{r linux-fix, eval=FALSE}
inla.binary.install()
```

## Common issues

- **Linux users**: you may need system libraries `libudunits2-dev`, `libgdal-dev`, `libgeos-dev`, `libproj-dev`, `libsqlite3-dev` and `libgsl-dev`.
- **Mac users**: make sure Xcode command line tools are installed (`xcode-select --install`)
- **Windows users**: Rtools may be required for some packages
